### 2024-07-16 journal/correctness.tex
cleaning up some .tex files (removing comments, \rev{})

A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
when the block is created.
% at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
Moreover, only a \op{CAS} at line \ref{setSuper1}  modifies  \fld{super}  
(from \nl\ to a non-\nl\ value), so it is changed only once.
Similarly, only a \op{CAS} at line \ref{cas} modifies an element of a node's \fld{blocks} array 
(from \nl\ to a non-\nl\ value), so blocks are permanently added to nodes.
Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
which implies the following.

@@

The following two lemmas show that no operation appears in more than one block of the root.
\begin{lemma} \label{lem::headProgress}
 If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$, then
 \begin{eqnarray*}
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{left}} \mbox{ and}\\
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{right}}.
 \end{eqnarray*}
\end{lemma}
\begin{proof}
Let $B$ be the block in $\var{v}.\fld{blocks}[b]$.
Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
read $b$ from $\var{v}.\head$ at line \ref{readHead}.
At that \rev{read}, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
created $B$ started.
It follows from Observation \ref{nonDecreasingHead} that the value that 
line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b$) 
stores in $B.\fld{end\sub{left}}$.
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
%of these calls to \op{CreateBlock} 
satisfy the claim.
\end{proof}

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two \rev{different} blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two \rev{different} blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
\rev{Suppose $B$ and $B'$ are in the same node $v$.
By Lemma \ref{lem::headProgress}, the $\fld{end\sub{left}}$ values of blocks of $v$ are non-decreasing,
so the intervals of direct subblocks of different blocks in $v$ in $v$'s left child defined by (\ref{defsubblock}) are non-overlapping.
Similarly, the intervals of direct subblocks in $v$'s right child are non-overlapping.
In all cases, the direct subblocks of $B$ and $B'$} (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

@@

%\begin{definition}
%$n\nf{.blocks[}i\nf{]}$ is \emph{established} if $n\nf{.head}>i$. An operation is \emph{established} in node $n$ 
%if it is in an established block of $n$. $EST^t_n$ is the set of established operations in node $n$ at time $t$.
%\end{definition}
%
%Now we want to say that blocks of a node grow over time.
%\begin{observation}\label{lem::blocksOrder}
%  If  time $t<$ time $t^\prime$ ($t$ is before $t^\prime$), then $ops(n.blocks)$ at time $t$ is a subset of 
%$ops(n.blocks)$ at time $t^\prime$.
%\end{observation}
%\begin{proof}
%Blocks are only appended (not modified) with \nf{CAS} to $n\nf{.blocks[}n\nf{.head]}$, so the set of the blocks of a node after the \nf{CAS} contains the set of the blocks before the \nf{CAS}.
%\end{proof}

% \begin{corollary}\label{lem::establishedOrder}
%   If  time $t<$ time $t^\prime$, then $EST_n^t\subseteq EST_n^{t^\prime}$.
% \end{corollary}
% \begin{proof}
% From Observations \ref{nonDecreasingHead}, \ref{lem::blocksOrder}.  
% \end{proof}

The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
on lines \ref{enqNew}, \ref{deqNew} %, \ref{createSumEnq} 
and \ref{createSumDeq} follows easily
from the definition of subblocks.

@@

Now, suppose $\var{v}$ is an internal node. By the definition of \rev{direct} subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the \rev{direct}
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the \enqueues\ in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and %those in 
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these \enqueues\ is 
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

In the next two lemmas, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
We say that node $\var{v}$ \emph{contains} an operation $op$ if some block in $\var{v}.\fld{blocks}$ contains $op$.
%\here{move this defn earlier and just recall it here?}
Since blocks are permanently added to nodes, if $\var{v}$ contains $op$ at some time, $v$ contains $op$ at all later times too.

@@

%\Eric{I think this case was missing from the proof in the thesis}
Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
&&\var{num\sub{enq}}\\
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of \enqueues\  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..$ $B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

In this section, we show that the routines used to compute responses to \op{Dequeue}
operations work correctly.
We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  We do this by showing that \rev{the value written in the \fld{super} field of a block $B$ in node $\var{v}$ is read from
\var{v}.\fld{parent}'s \var{head} field after $B$ is installed, but before $B_s$ is installed.}
%On one hand, $B.super$ is written before $B_s$ is installed:
%\op{Advance} writes $B.super$ before advancing $\var{v}.\head$  past $B$'s index, which must happen before the \op{CreateBlock} that creates
%$B_s$ gets the value of $B_s.\eleft$ or $B_s.\eright$.
%On the other hand, $B.\fld{super}$ cannot be written too long before $B_s$ is installed:
%$B.\fld{super}$ is written after $B$ is installed, and Lemma \ref{successfulRefresh} ensures that $B$ is propagated to the
%parent soon after.

\begin{lemma}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of \rev{direct} subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft} %or \ref{createEndRight} 
obtains a value greater than $b$.
By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least~1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ \rev{is assigned a non-\nl\ value before $B$'s superblock is installed}.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

%The reader may wonder when the case $b\nf{.super}=s$ happens. This can happen when $
%\nf{$n$.parent.blocks[$B$.super]}=\nf{null}$ when $B$\nf{.super} is written and $R_p$ puts its created block 
%into \nf{$n$.parent.blocks[$B$\nf{.super}]} afterwards.
To show  \op{GetEnqueue} and \op{IndexDequeue} work correctly, we  just  check that they correctly compute the index of the required block
and the operation's rank within the block.  
For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.
%\here{mention somewhere why precondition of IndexDequeue is true? also check that block of parent indexed by sup in that routine is non-null}

@@

Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
Since $B$ has been propagated to the root, its superblock $B'$ \rev{has also been propagated to the root}.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required \dequeue\ in the root.\end{proof}

@@

We \rev{next show in Lemma \ref{linearSat}} that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the concurrent execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  \rev{Then, we show in Lemma \ref{linearCorrect}} that each completed \dequeue\ returns the same result \rev{in the concurrent execution} as it would \rev{if the operations were performed sequentially in the order given by $L$}.

\begin{lemma} \label{linearSat}
$L$ is a legal linearization ordering.
\end{lemma}
\begin{proof}
By Corollary \ref{cor::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
By Lemma \ref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
so it appears in $L$.
\rev{Suppose an operation} $op_{1}$ terminates before another operation $op_{2}$ begins. By Lemma \ref{lem::appendExactlyOnce}, $op_{1}$ 
is propagated to the root before $op_2$ begins, \rev{and therefore before $op_2$ is propagated to the root.
Thus, $op_1$ appears in an earlier block of \var{root}.\fld{blocks} than $op_2$.
So, by the definition of $L$ in (\ref{linearization}), $op_1$ precedes $op_2$ in $L$.}
\end{proof}

@@

\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
\rev{computed on line \ref{computeNumEnq} and \ref{computeNumDeq}}
are the number of \enqueues\ and \dequeues\ contained in $B$, respectively.
Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$, \rev{as described in Equation (\ref{size-eqn})}.
\end{proof}

Next, we show each operation returns the same response as it would \rev{if the operations were performed sequentially in the order $L$}.

>>>>

A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
Moreover, only a \op{CAS} at line \ref{setSuper1}  modifies  \fld{super}  
(from \nl\ to a non-\nl\ value), so it is changed only once.
Similarly, only a \op{CAS} at line \ref{cas} modifies an element of a node's \fld{blocks} array 
(from \nl\ to a non-\nl\ value), so blocks are permanently added to nodes.
Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
which implies the following.

@@

The following two lemmas show that no operation appears in more than one block of the root.
\begin{lemma} \label{lem::headProgress}
 If $b>0$ and $\var{v}.\fld{blocks}[b] \neq \nl$, then
 \begin{eqnarray*}
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{left}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{left}} \mbox{ and}\\
 \var{v}.\fld{blocks}[b-1].\fld{end\sub{right}} &\leq& \var{v}.\fld{blocks}[b].\fld{end\sub{right}}.
 \end{eqnarray*}
\end{lemma}
\begin{proof}
Let $B$ be the block in $\var{v}.\fld{blocks}[b]$.
Before creating $B$ at line \ref{invokeCreateBlock}, the \op{Refresh} that installed $B$
read $b$ from $\var{v}.\head$ at line \ref{readHead}.
At that read, $\var{v}.\fld{blocks}[b-1]$ contained a block $B'$, by Invariant \ref{inv::headPosition}.
Thus, the \op{CreateBlock}($\var{v},b-1$) that created $B'$ terminated before the \op{CreateBlock}($\var{v},b$) that
created $B$ started.
It follows from Observation \ref{nonDecreasingHead} that the value that 
line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b-1$) stores in $B'.\fld{end\sub{left}}$   
is less than or equal to the value that line \ref{createEndLeft} of \op{CreateBlock}($\var{v},b$) 
stores in $B.\fld{end\sub{left}}$.
Similarly, the values stored in $B'.\eright$ and $B.\eright$ at line \ref{createEndRight} 
of these calls to \op{CreateBlock} 
satisfy the claim.
\end{proof}

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two different blocks in nodes at the same depth in the ordering tree, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two different blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
Suppose $B$ and $B'$ are in the same node $v$.
By Lemma \ref{lem::headProgress}, the $\fld{end\sub{left}}$ values of blocks of $v$ are non-decreasing,
so the intervals of direct subblocks of different blocks in $v$ in $v$'s left child defined by (\ref{defsubblock}) are non-overlapping.
Similarly, the intervals of direct subblocks in $v$'s right child are non-overlapping.
In all cases, the direct subblocks of $B$ and $B'$ (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

@@

The accuracy of the values stored in the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields
on lines \ref{enqNew}, \ref{deqNew}
and \ref{createSumDeq} follows easily
from the definition of subblocks.

@@

Now, suppose $\var{v}$ is an internal node. By the definition of direct subblocks in (\ref{defsubblock}) and Lemma \ref{lem::headProgress}, the direct
subblocks of $\var{v}.\fld{blocks}[1..i]$ are $\var{v}.\fld{left.blocks}[1..B.\eleft]$ 
and $\var{v}.\fld{right.blocks}[1..B.\eright]$.
Thus, the \enqueues\ in $E(\var{v}.\fld{blocks}[0])\cdots E(\var{v}.\fld{blocks}[i])$ are those in
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eleft])$ and %those in 
$E(\var{v}.\fld{left.blocks}[0]) \cdots E(\var{v}.\fld{left.blocks}[B.\eright])$.
By the hypothesis, the total number of these \enqueues\ is 
$\var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}$, 
which is the value that line \ref{createSumEnq} stored in $B.\fld{sum\sub{enq}}$ when $B$ was created.
The proof for \fld{sum\sub{deq}} (stored on line~\ref{createSumDeq}) is similar.
\end{proof}

@@

In the next two lemmas, we show two \op{Refresh}es suffice to propagate operations from a child to its parent.
We say that node $\var{v}$ \emph{contains} an operation $op$ if some block in $\var{v}.\fld{blocks}$ contains $op$.
Since blocks are permanently added to nodes, if $\var{v}$ contains $op$ at some time, $v$ contains $op$ at all later times too.

@@

Now suppose $R$'s call to \op{CreateBlock} returns \nl, causing $R$ to terminate at line \ref{addOP}.
Intuitively, since there are no operations in $\var{v}$'s children to promote, $op$ is already in $\var{v}$.
We formalize this intuition.
The value computed at line \ref{computeNumEnq} is
\begin{eqnarray*}
&&\var{num\sub{enq}}\\
&=&  \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}}\\
&&  - \var{v}.\fld{blocks}[h-1].\fld{sum\sub{enq}} \\
&=& \var{v}.\fld{left.blocks}[B.\eleft].\fld{sum\sub{enq}}\\
&& + \var{v}.\fld{right.blocks}[B.\eright].\fld{sum\sub{enq}} \\
&&- \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft].\fld{sum\sub{enq}} \\
&& - \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
It follows from Invariant \ref{inv::sum} that $num\sub{enq}$ is the total  number of \enqueues\  in 
$\var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[h-1].\eleft+1..B.\eleft]$ and
$\var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[h-1].\eright+1..$ $B.\eright]$.
Similarly, $num\sub{deq}$ is the total number of \dequeues\ contained in these blocks.
Since $num\sub{enq}+num\sub{deq}=0$ at line \ref{testEmpty},
these blocks contain no operations.
By Corollary \ref{blockNotEmpty}, this means the ranges of blocks are empty, so that $\var{v}.\fld{blocks}[h-1].\eleft \geq \var{B}.\eleft \geq i$.
Hence, $B_\ell$ is already a subblock of some block in $\var{v}$, so $\var{v}$ contains $op$.
\end{proof}

@@

In this section, we show that the routines used to compute responses to \op{Dequeue}
operations work correctly.
We first prove the \fld{super} field is accurate, since \op{IndexDequeue} uses it to trace superblocks up the tree.  We do this by showing that the value written in the \fld{super} field of a block $B$ in node $\var{v}$ is read from
\var{v}.\fld{parent}'s \var{head} field after $B$ is installed, but before $B_s$ is installed.

\begin{lemma}
\label{superRelation}
Let $B=\var{v}.\var{blocks}[b]$.
If $\var{v}.\fld{parent.blocks}[s]$ is the superblock of $B$ then $s-1\leq B.\fld{super}\leq s$.
\end{lemma}
\begin{proof}
We first show that $B.\fld{super}\leq s$.
Let $R_s$ be the instance of \op{Refresh}($\var{v}.\fld{parent}$) that installs $B$'s superblock 
in $\var{v}.\fld{parent.blocks}[s]$.
By the definition of direct subblocks (\ref{defsubblock}), $R_s$'s read $r$ of $\var{v}.head$ at line \ref{createEndLeft}
obtains a value greater than $b$.
By Invariant \ref{inv::headPosition}, $B.\fld{super}$ is not $\nl$ when $r$ occurs, which means
that $B.\fld{super}$ was set (by line~\ref{setSuper1}) to a value read from $\var{v}.\fld{parent}.\head$ before $r$.
When $r$ occurs, $\var{v}.\fld{parent.blocks}[s] = \nl$, since the later \op{CAS} by $R_s$ at line
\ref{cas} succeeds.
So, by Invariant \ref{inv::headPosition}, $\var{v}.\fld{parent}.\head \leq s$ when $r$ occurs.
Since the value stored in $B.\fld{super}$ was read from $\var{v}.\fld{parent.head}$ before $r$ and the \head\ field is non-decreasing by Observation \ref{nonDecreasingHead}, it follows that $B.super\leq s$.

Next, we show that $B.\fld{super}\geq s-1$.
The value stored in $B.\fld{super}$ at line \ref{setSuper1} is read from $\var{v}.\fld{parent}.\head$ at line \ref{readParentHead} and \head\ is always at least~1, so $B.\fld{super} \geq 1$.
So, if $s\leq 2$, the claim is trivial.  Assume $s>2$ for the rest of the proof.
By Lemma~\ref{lem::headProgress}, $\var{v}.\fld{parent.blocks}[s-1]$ 
is assigned a non-\nl\ value before $B$'s superblock is installed.  Let $R_{s-1}$ be the call to
$\op{Refresh}(\var{v}.\fld{parent})$ that installed the block in $\var{v}.\fld{parent.blocks}[s-1]$.
Let $r'$ be the step when $R_{s-1}$ reads $s-1$ in $\var{v}.\fld{parent}.\head$ at line \ref{readHead}.
This read $r'$ must be before $B$ is installed in $\var{v}$;
otherwise, Lemma \ref{successfulRefresh} would imply that $B$ is a subblock of one of 
$\var{v}.\fld{parent.blocks}[1..s-1]$, contrary to the hypothesis.
Now, consider the call to \op{Advance}($\var{v}, b$) that writes $B.\fld{super}$.
It is invoked either 
at line \ref{helpAdvance} after seeing $\var{v}.\fld{blocks}[b]\neq \nl$ at line \ref{ifHeadnotNull}
or at line \ref{advance} after ensuring $\var{v}.\fld{blocks}[b]\neq \nl$ at line~\ref{cas}.
Either way, the \op{Advance} is invoked after $B$ is installed, and therefore after $r'$.
By Observation \ref{nonDecreasingHead}, $\var{v}.\fld{parent}.\head$ is non-decreasing, so 
the value this \op{Advance} reads in $\var{v}.\fld{parent}.\head$ and
writes in $B.\fld{super}$ is greater than or equal to the value $s-1$ that $r'$ reads in $\var{v}.\fld{parent}.\head$.
\end{proof}

To show  \op{GetEnqueue} and \op{IndexDequeue} work correctly, we  just  check that they correctly compute the index of the required block
and the operation's rank within the block.  
For \op{IndexDequeue}, we use Lemma \ref{superRelation} each time \op{IndexDequeue} goes one step up the tree.

@@

Finally, \op{IndexDequeue} is called recursively on $\var{v}$'s parent.
Since $B$ has been propagated to the root, its superblock $B'$ has also been propagated to the root.
Thus, all preconditions of the recursive call are met.
By the induction hypothesis, the recursive call returns the location of the required \dequeue\ in the root.\end{proof}

@@

We next show in Lemma \ref{linearSat} that the linearization ordering $L$ defined in Equation
(\ref{linearization}) is a legal permutation of a subset of the operations in 
the concurrent execution, i.e., that it includes all operations that terminate and 
if one operation $op_1$ terminates before another operation $op_2$ begins, then $op_2$ does not precede $op_1$ in $L$.  Then, we show in Lemma \ref{linearCorrect} that each completed \dequeue\ returns the same result in the concurrent execution as it would if the operations were performed sequentially in the order given by $L$.

\begin{lemma} \label{linearSat}
$L$ is a legal linearization ordering.
\end{lemma}
\begin{proof}
By Corollary \ref{cor::noDuplicates}, $L$ is a permutation of a subset of the operations in the execution.
By Lemma \ref{lem::appendExactlyOnce}, each terminating operation is propagated to the root before it terminates,
so it appears in $L$.
Suppose an operation $op_{1}$ terminates before another operation $op_{2}$ begins. By Lemma \ref{lem::appendExactlyOnce}, $op_{1}$ 
is propagated to the root before $op_2$ begins, and therefore before $op_2$ is propagated to the root.
Thus, $op_1$ appears in an earlier block of \var{root}.\fld{blocks} than $op_2$.
So, by the definition of $L$ in (\ref{linearization}), $op_1$ precedes $op_2$ in $L$.
\end{proof}

@@

\begin{proof}
We prove the claim by induction on $b$. 
The base case when ${b=0}$ is trivially true, since the queue is initially empty and 
$\var{root}.\fld{blocks}[0]$ contains an empty block whose \size\ field is $0$. 
Assuming the claim holds for $b-1$, we prove it for $b$.
The \fld{size} field of the block $B$ installed in $\var{root}.\fld{blocks}[b]$ is computed
at line \ref{computeLength} of a call to \op{CreateBlock}(\var{root, b}).
By the induction hypothesis, $root.\fld{blocks}[b-1].\size$ gives the size of the queue before the operations
of block $B$ are performed.
By Invariant \ref{inv::sum}, the values of \var{num\sub{enq}} and \var{num\sub{deq}}
computed on line \ref{computeNumEnq} and \ref{computeNumDeq}
are the number of \enqueues\ and \dequeues\ contained in $B$, respectively.
Hence, the size of the queue after the operations of $B$ are performed (with \enqueues\ before \dequeues\ as specified by~$L$)
is $\max(0, \var{root}.\fld{blocks}[b-1].\size + \var{num\sub{enq}} - \var{num\sub{deq}})$, as described in Equation (\ref{size-eqn}).
\end{proof}

Next, we show each operation returns the same response as it would if the operations were performed sequentially in the order $L$.

\begin{lemma}\label{linearCorrect}
Each terminating \dequeue\ returns the response it would if the operations
were performed sequentially in the order given by the linearization $L$.
\end{lemma}
\begin{proof}
If a \dequeue\ $Deq$ terminates, it is contained in some block in the root, by Lemma \ref{lem::appendExactlyOnce}.
By Lemma \ref{lem::indexDequeue}, $Deq$'s call to \op{IndexDequeue} on line~\ref{invokeIndexDequeue}
returns a pair $\langle b,i\rangle$ such that $Deq$ is the $i$th \dequeue\ in the block 
$B=\var{root}.\fld{blocks}[b]$.
$Deq$ then calls \op{FindResponse}($b,i$) on line \ref{deqRest}.
By Lemma~\ref{sizeCorrectness}, the queue contains $\var{root}.\fld{blocks}[b-1].\fld{size}$ elements
after the operations in $\var{root}.\fld{blocks}[1..b-1]$ are performed sequentially 
in the order given by $L$.
By Invariant \ref{inv::sum}, the value of \var{num\sub{enq}} computed on line \ref{FRNum}
is the number of \enqueues\ in $B$.
Since the \enqueues\ in block $B$ precede the \dequeues,
the queue is empty when the $i$th \dequeue\ of $B$ occurs if 
$\var{root}.\fld{blocks}[b-1].\fld{size} + \var{num\sub{enq}} < i$.
So $Deq$ returns \nl\ on line \ref{returnNull} if and only if it would do so in the sequential
execution defined by $L$.
Otherwise, the size of the queue after doing the operations in $\var{root}.\fld{blocks}[0..b-1]$
in the sequential execution defined by $L$ is $\var{root}.\fld{blocks}[b-1].\fld{sum\sub{enq}}$ minus
the number of non-\nl\ \dequeues\ in that prefix of $L$.
Hence, line \ref{computeE} sets $e$ to the rank of $Deq$ among all the non-\nl\ \dequeues\ in $L$.
Thus, in the sequential execution defined by $L$, $Deq$ returns the value enqueued by the $e$th \enqueue\ in $L$.
By Invariant \ref{inv::sum}, this \enqueue\ is the $i_e$th \enqueue\ in $E(\var{root}.\fld{blocks}[b_e])$, where
$b_e$ and $i_e$ are the values $Deq$ computes on line \ref{FRb} and \ref{FRi}.
By Lemma \ref{lem::get}, the call to \op{GetEnqueue} returns the argument of the required \enqueue.
\end{proof}
